Variational Monte Carlo on bosonic systems

### 2018-02-06 project_1/article/article.tex
Add abstract section
Fix non-interacting case

@@

\section{Theory}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        V_{\text{ext}}(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    trap. The two-body Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one body operator, $h$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}\nabla_i^2
        + V_{\text{ext}}(\vf{r}_i),
    \end{align}
    (we assume equal mass)
    and the two-body interaction operator, $w$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
    \end{align}
    where $a$ is the hard sphere of the particle. The trial wavefunction we will
    be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \bra{\vf{r}}\ket{\Psi_T}
        = \bra{\vf{r}}\ket{\Phi_T}
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k) \\
        &= \Biggl(
            \prod_i^N g(\alpha, \beta, \vf{r}_i)
        \Biggr)
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
        \label{eq:initial_trial_wavefunction}
    \end{align}
    where $\alpha$ and $\beta$ are variational parameters and
    \begin{align}
        \vf{r} = (\vf{r}_1, \vf{r}_2, \dots, \vf{r}_N, \alpha, \beta).
    \end{align}
    Here $g$ are the single particle wavefunctions given by
    \begin{align}
        g(\alpha, \beta, \vf{r}_i)
        = \exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr] \equiv \phi(\vf{r}_i),
    \end{align}
    and $\ket{\Phi_T}$ the \textit{Slater permanent} consisting of the $N$ first
    single particle wavefunctions, and the correlation wavefunction, $f$, given
    by
    \begin{align}
        f(a, \vf{r}_j, \vf{r}_k)
        &=
        \begin{cases}
            0 & |\vf{r}_j - \vf{r}_k| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_j - \vf{r}_k|}
            \Bigr) & |\vf{r}_j - \vf{r}_k| > a.
        \end{cases}
    \end{align}

    \subsection{Local energy}
        As the many-body wavefunction creates a very large configuration space,
        where much of the wavefunction is small, we use the Metropolis algorithm
        in order to move towards regions in configuration space with
        ``sensible'' values. We define the \textit{local energy}, $E_L{\vf{r}}$,
        by
        \begin{align}
            E_L(\vf{r})
            &= \frac{\bra{\vf{r}}H\ket{\Psi_T}}{\bra{\vf{r}}\ket{\Psi_T}}.
        \end{align}
        If $\ket{\Psi_T}$ is an exact eigenfunction of the Hamiltonian, $E_L$
        will be constant. The closer $\ket{\Psi_T}$ is to the exact wave
        function, the less variation in $E_L$ as a function of $\vf{r}$ we get.
        One of the most computationally intensive part of the VMC algorithm will
        be to compute $E_L$. We therefore find an analytical expression for
        $E_L$ in terms of the trial wavefunction.

        \subsubsection{Non-interacting harmonic oscillator}
            We start by finding an analytical expression for the local energy
            with a system where we set $a = 0$ and $\beta = 1$, i.e., a system
            of non-interacting harmonic oscillators.
                = \bra{\vf{r}}\ket{\Phi_T}
                = \prod_i^N \exp\bigl[
                    -\alpha |\vf{r}_i|^2
                \bigr],
            where $|\vf{r}_i| = r_i$. As $a = 0$ the interaction term,
            $w(\vf{r}_i, \vf{r}_j)$, vanishes and the Hamiltonian is given by
            (in the spherical case)
                H &= \sum_i^N h(\vf{r}_i)
                = \sum_i^N \biggl(
                    -\frac{\hbar^2}{2m}\nabla_i^2
                    + \half m \omega^2 r_i^2
                \biggr).
            Working in spherical coordinates the gradient and the Laplace
            operator (applied to a scalar function) becomes
                \nabla f
                &= \vf{e}_r\tpl{r}f
                + \vf{e}_{\theta}\frac{1}{r}\tpl{\theta}f
                + \vf{e}_{\phi}\frac{1}{r\sin(\theta)}\tpl{\phi}f, \\
                \nabla^2 f
                &= \frac{1}{r^2}\tpl{r}
                \bigl[
                    r^2\tpl{r}f
                \bigr]
                + \frac{1}{r^2\sin(\theta)}\tpl{\theta}
                \bigl[
                    \sin(\theta)\tpl{\theta}f
                \bigr] \nonumber\\
                + \frac{1}{r^2\sin^2(\theta)}\tpl{\phi}^2f,
            where $\vf{e}_r$ signifies the unit vector in the radial direction
            and we use the shorthand notation
                \tpl{r}f \equiv \dpd[]{f}{r}.
            As the trial wavefunction for $\beta = 1$ is purely radial we get
            the same expression in one, two and three dimensions by setting
                r^2 =
                \begin{cases}
                    x^2 & \text{one dimension}, \\
                    x^2 + y^2 & \text{two dimensions}, \\
                    x^2 + y^2 + z^2 & \text{three dimensions}.
                \end{cases}
            The gradient then becomes
                \nabla_i\bra{\vf{r}}\ket{\Psi_T}
                &= \vf{e}_{{r_i}}\tpl{{r_i}}\prod_j^N\exp\bigl[
                    -\alpha r_j^2
                -2\alpha r_i\prod_j^N\exp\bigl[
                    -\alpha r_j^2
                \bigr]\vf{e}_{{r_i}} \\
                -2\alpha r_i\bra{\vf{r}}\ket{\Psi_T}\vf{e}_{{r_i}}.
            The Laplacian is given by
                \nabla_i^2\bra{\vf{r}}\ket{\Psi_T}
                &= \frac{1}{r_i^2}\tpl{{r_i}}
                \biggl\{
                    r_i^2\tpl{{r_i}}
                    \prod_j^N\exp\bigl[
                        -\alpha r_j^2
                    \bigr]
                \biggr\} \\
                \frac{1}{r_i^2}\tpl{{r_i}}
                \biggl\{
                    -2 \alpha r_i^3
                    \prod_j^N\exp\bigl[
                        -\alpha r_j^2
                    \bigr]
                \biggr\} \\
                \frac{1}{r_i^2}
                \Bigl\{
                    -6 \alpha r_i^2
                    + 4\alpha^2 r_i^4
                \Bigr\}
                \prod_j^N\exp\bigl[
                    -\alpha r_j^2
                \bigr]
                \Bigl\{
                    -6 \alpha
                    + 4\alpha^2 r_i^2
                \Bigr\}
            This gives the analytical expression for the local energy to be
                &= \frac{\bra{\vf{r}}H\ket{\Psi_T}}{\bra{\vf{r}}\ket{\Psi_T}} \\
                &= \sum_i^N \Bigl(
                    4\alpha^2 r_i^2
                    + \half m \omega^2 r_i^2
                \Bigr) - 6\alpha.
            The \textit{drift force} of the system is given by
                \vf{F}_i
                \frac{2\nabla_i\bra{\vf{r}}\ket{\Psi_T}}
                {\bra{\vf{r}}\ket{\Psi_T}} \\
                &= -4\alpha r_i\vf{e}_{{r_i}}.

>>>>

\begin{abstract}
    Something very abstract and clever should go here.
\end{abstract}

@@

\section{Theory}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        V_{\text{ext}}(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    trap. The two-body Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one body operator, $h$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}\nabla_i^2
        + V_{\text{ext}}(\vf{r}_i),
    \end{align}
    (we assume equal mass)
    and the two-body interaction operator, $w$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
    \end{align}
    where $a$ is the hard sphere of the particle. The trial wavefunction,
    $\ket{\Psi_T}$, we will
    be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \Phi_T(\vf{r})
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k) \\
        &= \Biggl(
            \prod_i^N g(\alpha, \beta, \vf{r}_i)
        \Biggr)
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
        \label{eq:initial_trial_wavefunction}
    \end{align}
    where $\alpha$ and $\beta$ are variational parameters and
    \begin{align}
        \vf{r} = (\vf{r}_1, \vf{r}_2, \dots, \vf{r}_N, \alpha, \beta).
    \end{align}
    Here $g$ are the single particle wavefunctions given by
    \begin{align}
        g(\alpha, \beta, \vf{r}_i)
        = \exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr] \equiv \phi(\vf{r}_i),
    \end{align}
    and $\ket{\Phi_T}$ the \textit{Slater permanent} consisting of the $N$ first
    single particle wavefunctions, and the correlation wavefunction, $f$, given
    by
    \begin{align}
        f(a, \vf{r}_j, \vf{r}_k)
        &=
        \begin{cases}
            0 & |\vf{r}_j - \vf{r}_k| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_j - \vf{r}_k|}
            \Bigr) & |\vf{r}_j - \vf{r}_k| > a.
        \end{cases}
    \end{align}

    \subsection{Local energy}
        As the many-body wavefunction creates a very large configuration space,
        where much of the wavefunction is small, we use the Metropolis algorithm
        in order to move towards regions in configuration space with
        ``sensible'' values. We define the \textit{local energy}, $E_L{\vf{r}}$,
        by
        \begin{align}
            E_L(\vf{r})
            &= \frac{H\Psi_T(\vf{r})}{\Psi_T(\vf{r})}.
        \end{align}
        If $\ket{\Psi_T}$ is an exact eigenfunction of the Hamiltonian, $E_L$
        will be constant. The closer $\ket{\Psi_T}$ is to the exact wave
        function, the less variation in $E_L$ as a function of $\vf{r}$ we get.
        One of the most computationally intensive parts of the VMC algorithm
        will be to compute $E_L$. We therefore find an analytical expression for
        $E_L$ in terms of the trial wavefunction.

        \subsubsection{Non-interacting harmonic oscillator}
            We start by finding an analytical expression for the local energy in
            a system where we set $a = 0$ and $\beta = 1$, i.e., a system of
            non-interacting harmonic oscillators.
            \begin{align}
                \Psi_T(\vf{r})
                = \Phi_T(\vf{r})
                = \prod_i^N \exp\bigl[
                    -\alpha |\vf{r}_i|^2
                \bigr],
            \end{align}
            where $|\vf{r}_i| = r_i$. As $a = 0$ the interaction term,
            $w(\vf{r}_i, \vf{r}_j)$, vanishes and the Hamiltonian is given by
            (in the spherical case)
            \begin{align}
                H &= \sum_i^N h(\vf{r}_i)
                = \sum_i^N \Biggl(
                    -\frac{\hbar^2}{2m}\nabla_i^2
                    + \half m \omega^2 r_i^2
                \Biggr).
            \end{align}
            The hardest part of finding the local energy is computing the
            kinetic term. Starting with the gradient of the wavefunction in the
            non-interacting case, we get
            \begin{align}
                \nabla_k\Psi_T(\vf{r})
                &= -2\alpha \vf{r}_k\Psi_T(\vf{r}).
            \end{align}
            The Laplacian yields
            \begin{align}
                \nabla^2_k\Psi_T(\vf{r})
                &= \big(-2d\alpha + 4\alpha^2 r_k^2\bigr)\Psi_T(\vf{r}),
            \end{align}
            where $d$ is the dimensionality of the problem determined by
            $\vf{r}_k \in \mathbb{R}^d$. This yields the analytical expression
            for the local energy in the non-interacting case to be
            \begin{align}
                E_L(\vf{r})
                &=
                \sum_i^N
                \Biggl(
                    -\frac{\hbar^2}{2m}
                    \bigl[
                        -2d\alpha + 4\alpha^2 r_i^2
                    \bigr]
                    + \half m\omega^2 r_i^2
                \Biggr).
            \end{align}
            In natural units, with $\hbar = c = m = 1$, this reduces to
            \begin{align}
                E_L(\vf{r})
                &=
                \alpha dN
                + \biggl(
                    \half\omega^2
                    - 2\alpha^2
                \biggr)
                \sum_i^N r_i^2.
            \end{align}
            It is worth noting that for $\omega = 1$ we will find a stable value
            for the local energy when $\alpha = \half$. This happens as the
            entire sum over all the random walkers disappears.

%            Working in spherical coordinates the gradient and the Laplace
%            operator (applied to a scalar function) becomes
%            \begin{align}
%                \nabla f
%                &= \vf{e}_r\tpl{r}f
%                + \vf{e}_{\theta}\frac{1}{r}\tpl{\theta}f
%                + \vf{e}_{\phi}\frac{1}{r\sin(\theta)}\tpl{\phi}f, \\
%                \nabla^2 f
%                &= \frac{1}{r^2}\tpl{r}
%                \bigl[
%                    r^2\tpl{r}f
%                \bigr]
%                + \frac{1}{r^2\sin(\theta)}\tpl{\theta}
%                \bigl[
%                    \sin(\theta)\tpl{\theta}f
%                \bigr] \nonumber\\
%                &\qquad
%                + \frac{1}{r^2\sin^2(\theta)}\tpl{\phi}^2f,
%            \end{align}
%            where $\vf{e}_r$ signifies the unit vector in the radial direction
%            and we use the shorthand notation
%            \begin{align}
%                \tpl{r}f \equiv \dpd[]{f}{r}.
%            \end{align}
%            As the trial wavefunction for $\beta = 1$ is purely radial we get
%            the same expression in one, two and three dimensions by setting
%            \begin{align}
%                r^2 =
%                \begin{cases}
%                    x^2 & \text{one dimension}, \\
%                    x^2 + y^2 & \text{two dimensions}, \\
%                    x^2 + y^2 + z^2 & \text{three dimensions}.
%                \end{cases}
%            \end{align}
%            The gradient then becomes
%            \begin{align}
%                \nabla_i\bra{\vf{r}}\ket{\Psi_T}
%                &= \vf{e}_{{r_i}}\tpl{{r_i}}\prod_j^N\exp\bigl[
%                    -\alpha r_j^2
%                \bigr] \\
%                &=
%                -2\alpha r_i\prod_j^N\exp\bigl[
%                    -\alpha r_j^2
%                \bigr]\vf{e}_{{r_i}} \\
%                &=
%                -2\alpha r_i\bra{\vf{r}}\ket{\Psi_T}\vf{e}_{{r_i}}.
%            \end{align}
%            The Laplacian is given by
%            \begin{align}
%                \nabla_i^2\bra{\vf{r}}\ket{\Psi_T}
%                &= \frac{1}{r_i^2}\tpl{{r_i}}
%                \biggl\{
%                    r_i^2\tpl{{r_i}}
%                    \prod_j^N\exp\bigl[
%                        -\alpha r_j^2
%                    \bigr]
%                \biggr\} \\
%                &=
%                \frac{1}{r_i^2}\tpl{{r_i}}
%                \biggl\{
%                    -2 \alpha r_i^3
%                    \prod_j^N\exp\bigl[
%                        -\alpha r_j^2
%                    \bigr]
%                \biggr\} \\
%                &=
%                \frac{1}{r_i^2}
%                \Bigl\{
%                    -6 \alpha r_i^2
%                    + 4\alpha^2 r_i^4
%                \Bigr\}
%                \prod_j^N\exp\bigl[
%                    -\alpha r_j^2
%                \bigr]
%                \\
%                &=
%                \Bigl\{
%                    -6 \alpha
%                    + 4\alpha^2 r_i^2
%                \Bigr\}
%                \bra{\vf{r}}\ket{\Psi_T}.
%            \end{align}
%            This gives the analytical expression for the local energy to be
%            \begin{align}
%                E_L(\vf{r})
%                &= \frac{\bra{\vf{r}}H\ket{\Psi_T}}{\bra{\vf{r}}\ket{\Psi_T}} \\
%                &= \sum_i^N \Bigl(
%                    4\alpha^2 r_i^2
%                    + \half m \omega^2 r_i^2
%                \Bigr) - 6\alpha.
%            \end{align}
%            The \textit{drift force} of the system is given by
%            \begin{align}
%                \vf{F}_i
%                &=
%                \frac{2\nabla_i\bra{\vf{r}}\ket{\Psi_T}}
%                {\bra{\vf{r}}\ket{\Psi_T}} \\
%                &= -4\alpha r_i\vf{e}_{{r_i}}.
%            \end{align}
